---
title: "week4_3assumption_NOT_met"
author: "Hippo"
date: "`r format(Sys.time(), '%Y_%m_%d_%H')`"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    fig_width: 4
    fig_height: 3
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
library(tidyverse)
library(ggplot2)
```

### get data
```{r}
library(survival)
library(car)
course_dir <- "/Users/minghan/Google Drive/Coursera_survival_analysis"
g <- read.csv(file.path(course_dir, "GMPH_1K_final_2.csv"), header = TRUE, sep = ",")
g %>% dim()
# g %>% head()
g$gender <- as.factor(g$gender)
g$copd <- as.factor(g$copd)
g$ethnicgroup <- as.factor(g$ethnicgroup)
levels(g$ethnicgroup) <- c(levels(g$ethnicgroup), "8")
g$ethnicgroup[is.na(g$ethnicgroup)] <- "8"
table(g$ethnicgroup, exclude = NULL)

# set quintile 0 to NAs, coxph will ignore these patients
g <- g %>% mutate(quintile = recode(quintile, "c('0')=NA"))
table(g$quintile, exclude = NULL)
g$quintile <- as.factor(g$quintile)
```

### what does NON-PH mean?
- example: gender
    - if relation btw male vs female risk of death changes over time,
        - it could mean that males have higher risk of death early on,
        - then relation changes, and females have higher risk of death later on
    - there is **interaction** btw gender and time
        - { can you have interaction btw variable and OUTCOME??? }
    - all you need to do is to add an interaction term btw gender and time {...}
        - { need to find a dataset with actual interaction,
        - and plot out KM curves with and without interaction and see for myself... }
- testing if interaction term is statistically significant is another way of testing PH assumption
    - if interaction term not stat sig, then PH assumption is valid
    - should get p-value and plots
- for mathematically reasons, cannot just include follow-up time itself,
    - need to transform it with tt() time-transform function
```{r}
# model with just gender
cox <- coxph(Surv(fu_time, death) ~ gender, data = g)
summary(cox)

# model with gender and gender * fu_time
cox_int <- coxph(Surv(fu_time, death) ~ gender + tt(gender), data = g)
summary(cox_int)
```

- p-value of 0.529 for tt(gender) is not statistically significant, good news, 
- PH assumption is valid, no interaction between variable and time
- i.e. time does NOT affect gender's impact on risk of death

### scenario 2
- if p-value is low, but hazards looks proportional for most the the follow-up period
- can divide survival analysis into **2 time periods**
    - fit one model without interaction for when PH is valid
    - fit another model with interaction for when PH is invalid

### scenario 3
- stratify by variable
- e.g. separate models for male and females
- disadvantage = can no longer assess effect of gender on mortality

### EOF