---
title: "week3_1EDA_multivariateCox"
author: "Hippo"
date: "`r format(Sys.time(), '%Y_%m_%d_%H')`"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    fig_width: 4
    fig_height: 3
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
library(tidyverse)
library(ggplot2)
```

### intro
- multiple predictors have their own issues {collinearity?}
- run descriptive stats on age, gender, prior_dnas, ethnic group, COPD

```{r, results="hold"}
library(survival)
course_dir <- "/Users/minghan/Google Drive/Coursera_survival_analysis"
g <- read.csv(file.path(course_dir, "GMPH_1K_final_2.csv"), header = TRUE, sep = ",")
g %>% dim()
# g %>% head()
g$gender <- as.factor(g$gender)
g$prior_dnas <- as.factor(g$prior_dnas)
g$ethnicgroup <- as.factor(g$ethnicgroup)
g$copd <- as.factor(g$copd)
```
#### age
```{r}
summary(g$age)
```

- nice continuous variable without any NAs

#### gender
```{r}
table(g$gender, exclude = NULL)
```

- nice binary variable with a lot of people in each category

#### copd
```{r}
table(g$copd, exclude = NULL)
```

- having copd is a lot less than not having copd
- there's usually under-recording of comorbidities for various reasons (e.g. not diagnosed)

#### ethnic group
```{r}
table(g$ethnicgroup, exclude = NULL)
levels(g$ethnicgroup) <- c(levels(g$ethnicgroup),"8")
g$ethnicgroup[is.na(g$ethnicgroup)] <- "8"
table(g$ethnicgroup, exclude = NULL)
```

- turn NAs into another category so can include those patients in analysis

#### dealing with prior_dnas
```{r}
table(g$prior_dnas, exclude = NULL)
```

- option1 = pretend it's continuous, assume linear relationship with outcome
    - preferred for ordinal variables with lots of categories
    - but need to check relation is indeed linear with outcome
- option2 = use existing categories
    - categorising loses information
    - fitting categorical variable with small categories will cause issues (more later)
- option3 = combine some small categories
    - if relation is non-linear, and have small categories, then option3 is best
    - assessing linearity is made harder by the small categories too

### multivariate cox
```{r}
# try model with prior_dnas as continuous var
g$prior_dnas <- as.integer(g$prior_dnas)

cox_multi <- coxph(Surv(fu_time, death) ~ age + gender + copd + prior_dnas + ethnicgroup, data = g)
summary(cox_multi)
```

- gender1 = "male"
- ethnicity1 = "white"
- changing reference category covered later

#### interpretation
- **age**, 1.06411 (6%) increase in risk of mortality for every year you age
- **females** (gender2) has 0.77473 (22%) lower risk of mortality following hospital admission for heart failure vs males
- **prior_dnas** is a significant predictor of risk of mortality,
    - every 1 more missed visits = 1.18614 (18.6%) increase in risk
- **ethnicgroup9** has 1.4966 (49.7%) increase in risk compared to ethnicgroup1,
    - but results not significant, {probably because small group}
- each of the predictors is ADJUSTED for the other predictors {so results would change if you omitted some predictors}
<br><br>

### trying survfit() different prior_dnas groupings
- looking at width of CI for each category will tell you if you have enough patients in each to get significant results
```{r}
library(car) # need recode from car package, not from dplyr
table(g$prior_dnas, exclude = NULL)
g$prior_dnas <- factor(g$prior_dnas)
g$prior_dnas %>% levels()
g <- g %>% mutate(prior_dnas4groups = recode(prior_dnas,"c('4','5','6','7','8','9','10')='4'"))
table(g$prior_dnas4groups, exclude = NULL)

cox_prior_dnas4groups <- coxph(Surv(fu_time, death) ~ age + gender + copd + prior_dnas4groups + ethnicgroup, data = g)
summary(cox_prior_dnas4groups)
```

- looks like prior_dnas4groups4 has 2.196314 risk compared to reference group


#### EOF