---
title: "week4_5practice_choosing_predictors"
author: "Hippo"
date: "`r format(Sys.time(), '%Y_%m_%d_%H')`"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    fig_width: 4
    fig_height: 3
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = TRUE)
library(tidyverse)
library(ggplot2)
```

### get data
```{r, results="hide"}
library(survival)
library(car)
course_dir <- "/Users/minghan/Google Drive/Coursera_survival_analysis"
g <- read.csv(file.path(course_dir, "GMPH_1K_final_2.csv"), header = TRUE, sep = ",")
g %>% dim()
# g %>% head()
g$gender <- as.factor(g$gender)
g$copd <- as.factor(g$copd)
g$ethnicgroup <- as.factor(g$ethnicgroup)
levels(g$ethnicgroup) <- c(levels(g$ethnicgroup), "8")
g$ethnicgroup[is.na(g$ethnicgroup)] <- "8"
table(g$ethnicgroup, exclude = NULL)

# set quintile 0 to NAs, coxph will ignore these patients
g <- g %>% mutate(quintile = recode(quintile, "c('0')=NA"))
table(g$quintile, exclude = NULL)
g$quintile <- as.factor(g$quintile)
g %>% names()
```

### option1, using apriori variables
- https://www.ncbi.nlm.nih.gov/books/NBK513479/  
- Objective 1: what are the main patient, primary care and hospital factors associated with variation in readmission and mortality rates?
- Table 6: Odds ratios with 95% CIs for possible patient, trust and primary care predictors of mortality within 1 year
of admission for HF and COPD patients
- paper used logistic regression with outcome = death within 1 year of follow-up, but interesting to see if significant predictors apply to Cox regression
- significant predictors = age, gender (male = ref), ethnicgroup (white = ref), ihd, stroke, valvular_disease, pvd, copd, pneumonia, hypertension, cancer, metastatic_cancer, senile, dementia, mental health, prior_dnas, los (0 nights = ref), cabg
<br><br>
- use age, los and prior_dnas as continuous for this exercise

```{r, results="hide"}
# check class of each column
g %>% dim() # [1] 1000   31
g %>% str()

# get new df for selected variables
g1 <- g %>% select(c(age, gender, ethnicgroup, ihd, stroke, valvular_disease, pvd, copd, pneumonia, hypertension, cancer, metastatic_cancer, senile, dementia, mental_health, prior_dnas, los, cabg, fu_time))
g1 %>% dim() # [1] 1000   21

# convert all categorical variables to factors
g1 <- g1 %>% mutate_if(~ n_distinct(.) == 2, as.factor)
g1$ethnicgroup <- as.factor(g1$ethnicgroup)

# attach death on its own, since previous mutate_if() affects it
g1$death <- g$death
g1 %>% str()

# run cox model on apriori variables
cox <- coxph(Surv(fu_time, death) ~ ., data = g1)
summary(cox)
```

### option2, include all available predictors
```{r, results="hide"}
g %>% str()
g2 <- g %>% select(-c(id, death))
g2 <- g2 %>% mutate_if(~n_distinct(.) == 2, as.factor)
g2$death <- g$death
g2 %>% str()
g2$prior_dnas %>% table(., exclude = NULL)
g2$ethnicgroup %>% table(., exclude = NULL)
g2$quintile %>% table(., exclude = NULL)

# only 3 patients had crt=1, model could not converge, remove crt
g2$crt %>% table(., exclude = NULL)
g2 <- g2 %>% select(-crt)

# fit model2 with all available predictors
cox2 <- coxph(Surv(fu_time, death) ~ ., data = g2)
cox2_coef_before <- summary(cox2)$coefficients %>%
  as.data.frame() %>%
  rownames_to_column(., "predictors") %>%
  rename("HR_before" = `exp(coef)`)
cox2_coef_before
```

### option3, backward elimination
```{r}
# get significant variables
vars_sig <- summary(cox2)$coefficients %>% 
  as.data.frame() %>%
  rownames_to_column(., "predictors") %>% 
  filter(`Pr(>|z|)` < 0.05) %>%
  select(predictors) %>%
  unlist() %>% unname()
vars_sig <- vars_sig %>% gsub('[0-9]+', '', .)

# rerun model with only significant variables
fmla <- as.formula(paste("Surv(fu_time, death) ~ ", paste(vars_sig, collapse=" + "), sep = ""))
cox2_vars_sig <- coxph(fmla, data = g2)
cox2_coef_after <- summary(cox2_vars_sig)$coefficients %>%
  as.data.frame() %>% 
  rownames_to_column(., "predictors") %>%
  rename("HR_after" = `exp(coef)`)
cox2_coef_after

# compare cox2_before vs cox2_after HRs, see which has > 0.05 difference
inner_join(cox2_coef_before, cox2_coef_after, by = "predictors") %>%
  select(predictors, HR_before, HR_after) %>%
  mutate(HR_diff = HR_after - HR_before) %>%
  filter(abs(HR_diff) > 0.05)

"          predictors HR_before  HR_after    HR_diff
1 metastatic_cancer1 8.3178427 10.314927 1.99708467
2       ethnicgroup8 0.9816789  1.065319 0.08364040
3       ethnicgroup9 1.4651009  1.533574 0.06847298"
```

```{r}
vars_sig
```


- looks like **metastatic_cancer** and **ethnicgroup**'s HR changed > 0.05

#### add non-sig variables back
```{r, results="hide"}
# get list of non-significant variables to add back
vars_sig
vars_all <- names(g2)
vars_to_addback <- setdiff(vars_all[!vars_all %in% c("fu_time", "death")], vars_sig); vars_to_addback

i = 1
for (i in 1:length(vars_to_addback)){
  bkwd_elim_vars <- c(vars_sig, vars_to_addback[i]); bkwd_elim_vars
  fmla <- as.formula(paste("Surv(fu_time, death) ~ ", paste(bkwd_elim_vars, collapse=" + "), sep = ""))
  cox2_bkwd_elim_vars <- coxph(fmla, data = g2)
  bkwd_elim_HRs <- summary(cox2_bkwd_elim_vars)$coefficients %>%
    as.data.frame() %>%
    rownames_to_column(., "predictors") %>%
    rename("HR_bkwd_elim" = `exp(coef)`) %>%
    select(predictors, HR_bkwd_elim) %>%
    filter(predictors %in% c("metastatic_cancer1", "ethnicgroup8", "ethnicgroup9"))
  print(paste0("added <", vars_to_addback[i], "> back:"))
  print(paste0("bkwd_elim_HRs are:"))
  print(bkwd_elim_HRs)
  print("")
}
# [1] "added <cancer> back:"
# [1] "bkwd_elim_HRs are:"
#           predictors HR_bkwd_elim
# 1 metastatic_cancer1     8.583697   <--
# 2       ethnicgroup8     1.077854
# 3       ethnicgroup9     1.551762

# [1] "added <renal_disease> back:"
# [1] "bkwd_elim_HRs are:"
#           predictors HR_bkwd_elim
# 1 metastatic_cancer1     9.895061   <--
# 2       ethnicgroup8     1.050404
# 3       ethnicgroup9     1.541174

# [1] "added <arrhythmias> back:"
# [1] "bkwd_elim_HRs are:"
#           predictors HR_bkwd_elim
# 1 metastatic_cancer1    10.566096
# 2       ethnicgroup8     1.027209
# 3       ethnicgroup9     1.478646   <--

# [1] "added <pneumonia> back:"
# [1] "bkwd_elim_HRs are:"
#           predictors HR_bkwd_elim
# 1 metastatic_cancer1    10.623099
# 2       ethnicgroup8     1.021901
# 3       ethnicgroup9     1.473842   <--
```

- looks like adding "cancer" variable back brought "metastatic_cancer" to HR=8.58 (original HR=8.32)
    - but ethnicgroup HR's did not change
- if bring "arrhythmias" or "pneumonia" back, "ethnicgroup9" HR goes to 1.47 (original HR=1.47)
    - but now "metastatic_cancer" HR=10.62 again
- so how to deal with this when multiple variable's HR changes after removing non-significant variables??
<br><br>
- seems like in final model, adding back "cancer", "pneumonia", "arrhythmias" will get both "metastatic_cancer" and "ethnicgroup" HR's back to original

### final model
```{r}
final_vars <- c(vars_sig, "cancer", "pneumonia", "arrhythmias"); final_vars
fmla <- as.formula(paste("Surv(fu_time, death) ~ ", paste(final_vars, collapse=" + "), sep = ""))
cox2_final_vars <- coxph(fmla, data = g2)
summary(cox2_final_vars)
```

### testing assumptions

#### 1-PH assumption
- **Schoenfeld residuals**

##### p-value
```{r}
zph <- cox.zph(cox2_final_vars, transform = "km", global = TRUE) 
# defaults to "km" and global = TRUE
print(zph)
plot(zph)
```

```{r, fig.height=10, fig.width=10}
library(survminer)
ggcoxzph(zph, var = final_vars,
         font.main = 10, 
         font.x = c(10, "plain", "black"), 
         font.y = c(10, "plain", "black"), 
         font.tickslab = c(8, "plain", "black"))
```

- <span style="color:Tomato">ihd had p-value of only 0.0032, should this var be removed?</span>
    - or should we rely on GLOBAL p-value?
<br><br>
- how come for some vars, there's more positive residuals than negatives?
    - e.g. los, dementia, valvular_disease, metastatic_cancer, prior_dnas, cancer
- only quintile and ethnicgroup had the opposite problem - more negative residuals than positive...

##### graphically with KM plot
- only works well for binary vars
```{r}
km_fit <- survfit(Surv(fu_time, death) ~ gender, data = g2)
ggsurvplot(km_fit, data = g2)
```

#### 2-outliers

##### influence of obs on HR (coefficients)
```{r, fig.height=10, fig.width=10}
ggcoxdiagnostics(cox2_final_vars, type = "dfbeta",
                 linear.predictions = FALSE,
                 ggtheme = theme_bw())
```

##### distribution of **deviance residuals**
```{r}
deviance_res <- residuals(cox2_final_vars, type = "deviance")
deviance_res_outside1.96SD <- deviance_res[abs(deviance_res) > 1.96]
deviance_res_outside1.96SD %>% length / deviance_res %>% length
# [1] 0.06161616
# not bad, should be < 0.05
```

```{r}
ggcoxdiagnostics(cox2_final_vars, type = "deviance",
                 linear.predictions = FALSE,
                 ggtheme = theme_bw())
```

#### 3-linear relation, continuous var <-> outcome
- **Martingale residuals**
```{r}
# g2 %>% select(final_vars) %>% str()
ggcoxfunctional(Surv(fu_time, death) ~ age + log(age) + sqrt(age), data = g2)
ggcoxfunctional(Surv(fu_time, death) ~ los + sqrt(los), data = g2) # can't log(los), it has some zeros, log(0) = -Inf
ggcoxfunctional(Surv(fu_time, death) ~ prior_dnas + sqrt(prior_dnas), data = g2) # can't log(prior_dnas) neither, same problem
```

- <span style="color:Tomato">why log() and sqrt() ???</span>
    - to see if after log() or sqrt() the relationship is more linear
- <span style="color:Tomato">what if you can't log() cuz there's 0's???</span>
- <span style="color:Tomato">what counts as linear relationship from these graphs...?</span>

---
### issues encountered
- choosing vars to include from the paper
- backward elimination
    - what if multiple significant var HR changed > 0.05? and adding back non-significant var to restore HR restores either of significant var but not all significant vars???
        - A: actually in final model if added back important non-significant vars, will correct their respective significant var's HRs
    - <span style="color:Tomato">but why does adding back non-significant vars affect significant var's HR in the first place???</span>
        - how are these variables "correlated"???
        - I thought we don't want correlated variables in the same model???
- choosing final set of predictors
- testing proportionality assumptions
    - ihd had p-value of only 0.0032, should this var be removed? or should we just rely on GLOBAL p-value?
    - how to interpret graph from Martingale residuals? what counts as linear relationship???
- why final set of predictors differ from paper
    - {logistic vs survival analysis; different dataset}

### EOF