---
title: "week4_2test_Coxmodel_assumptions.Rmd"
author: "Hippo"
date: "`r format(Sys.time(), '%Y_%m_%d_%H')`"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    fig_width: 4
    fig_height: 3
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
library(tidyverse)
library(ggplot2)
```

### get data
```{r}
library(survival)
library(car)
course_dir <- "/Users/minghan/Google Drive/Coursera_survival_analysis"
g <- read.csv(file.path(course_dir, "GMPH_1K_final_2.csv"), header = TRUE, sep = ",")
g %>% dim()
# g %>% head()
g$gender <- as.factor(g$gender)
g$copd <- as.factor(g$copd)
g$ethnicgroup <- as.factor(g$ethnicgroup)
levels(g$ethnicgroup) <- c(levels(g$ethnicgroup), "8")
g$ethnicgroup[is.na(g$ethnicgroup)] <- "8"
table(g$ethnicgroup, exclude = NULL)

# set quintile 0 to NAs, coxph will ignore these patients
g <- g %>% mutate(quintile = recode(quintile, "c('0')=NA"))
table(g$quintile, exclude = NULL)
g$quintile <- as.factor(g$quintile)
```

### test PH assumption
- e.g. test if hazard is proportional for a given predictor (e.g. gender)
- test graphically
    - when you plot the hazards for male vs female,
    - lines should be approximately parallel
    - at any time point, you can multiply hazard of male, get hazard of female
    - then the relationship can be summarized with 1 number, the HR
    - **hazard functions should definitely NOT cross**
- test with p-value
    - using **Schoenfeld** residuals
    - high p-value is good, means residuals not correlated with time of event, proportional assumption valid
- should do both graphically and with p-value

```{r, out.height="50%", out.width="50%"}
cox <- coxph(Surv(fu_time, death) ~ age + gender + copd + quintile + ethnicgroup, data = g)

zph <- cox.zph(cox, transform = "km", global = TRUE) # defaults to "km" and global = TRUE
print(zph)
plot(zph)

# run on gender only
cox_gender <- coxph(Surv(fu_time, death) ~ gender, data = g)

zph_gender <- cox.zph(cox_gender, transform = "km", global = TRUE) # defaults to "km" and global = TRUE
print(zph_gender)
plot(zph_gender)
# residual is different if run gender only, vs multivariate
```

- pvalue tests null model = if slope for regression line of residuals is 0
    - i.e. if slope = 0, accept null model, p-value > 0.05
- pvalue only testing for linear relationship btw residuals and time,
    - can miss non-linear relationships
    - should check plots too
- plots show **scaled** Schoenfeld residuals with time
<br><br>
- more later on what to do if PH assumption NOT met

#### survminer ggcoxzph()
- time is transformed...
```{r}
library(survminer)
ggcoxzph(zph_gender)
```

#### KM plot
- to visually check hazard functions
- only works well for binary variables (e.g. gender)
```{r, out.height="50%", out.width="50%"}
km_fit <- survfit(Surv(fu_time, death) ~ gender, data = g)
plot(km_fit, xlab = "time", ylab = "survival probability")
```

### deviance residuals
- normalized transformation of Martingale residuals
- mean 0, sd 1
- helps to check for outliers/ influential data points (e.g. patient)
- either
    - examine influence of each data point on *coefficient*, or
    - plot distribution of residuals against *covariate*

#### influence of each obs on regression coefficient
- using **survminer's ggcoxdiagnostics(... type = "dfbeta")**
- dfbeta plots estimated changes in regression coefficient on deleting each observation (patient) in turn
```{r, out.height="50%", out.width="50%"}
cox <- coxph(Surv(fu_time, death) ~ age, data = g)
ggcoxdiagnostics(cox, type = "dfbeta",
                 linear.predictions = FALSE,
                 ggtheme = theme_bw())
```

#### distribution of residuals against covariate
- residuals SHOULD BE normally distributed (mean 0, SD 1)
    - then only 5% of observations should be 1.96 SD from mean
    - if > 5% observations 1.96 SD from mean,
        - then model doesn't fit data as well, and
        - some observations are outliers
    - papers report "proportion of standardized residuals outside -1.96 to 1.96 range = e.g. 0.6%"
        - {shouldn't this be for each covariate? or for all covariates... 
        - nevermind, if multivariate then one residual per observation for all variables}
```{r, out.height="50%", out.width="50%"}
cox <- coxph(Surv(fu_time, death) ~ age, data = g)
ggcoxdiagnostics(cox, type = "deviance",
                 linear.predictions = FALSE,
                 ggtheme = theme_bw())
```

- residuals = E - O
- positive values - individuals expected/ predicted survival times > observed survival times
- negative values - individuals expected/ predicted survival times < observed survival times
- observations with very small or large residuals are outliers, poorly predicted by the model

### Martingale residuals, linear relation
- tests whether continuous variables have linear relation with outcome
- "if you fit age as a single term in the model..." {what if other terms were included in the model?}
```{r}
ggcoxfunctional(Surv(fu_time, death) ~ age + log(age) + sqrt(age), data = g)
# oh okay, age + log(age) + sqrt(age) just plots 3 separate graphs...
```

- Martingale residuals range from 1 to -inf
    - MR near 1 = patients Exp death > Obs death
    - MR large neg = patients Exp death < Obs death

### practice: test PH assumption for COPD
```{r}
cox_copd <- coxph(Surv(fu_time, death) ~ copd, data = g)
zph_copd <- cox.zph(cox_copd, transform = "km", global = TRUE) # defaults to "km" and global = TRUE
print(zph_copd)
```

### EOF