---
title: "week2_2runCoxModel_inR"
author: "Hippo"
date: "`r format(Sys.time(), '%Y_%m_%d_%H')`"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    fig_width: 4
    fig_height: 3
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
library(tidyverse)
library(ggplot2)
```

```{r, results="hold"}
library(survival)
course_dir <- "/Users/minghan/Google Drive/Coursera_survival_analysis"
g <- read.csv(file.path(course_dir, "GMPH_1K_final_2.csv"), header = TRUE, sep = ",")
g %>% dim()
g$fu_time %>% typeof()
g$death %>% typeof()
```

### run simple Cox model
```{r}
cox <- coxph(Surv(fu_time, death) ~ age, data = g)
summary(cox)
```

- coef = log hazard ratio
- exp(coef) = hazard ratio
    - HR of 1.06 means for every 1 year increase in age, 6% increase in risk of death
    - older people have higher risk of death vs younger people (as expected)
    - fitted age as continuous variable using linear model, so 6% HR is constant
- se is small (only 1 predictor used, lots of patients)
- p-value is tiny (highly statistically significant predictor of risk of death)
- concordance
    - similar to c-statistic in logistic regression
    - fraction of pairs of patients in a sample where patient with higher survival time was predicted to have higher risk of death (i.e. the model was correct)
    - values > 0.5 is better than chance
- R^2 (didn't output this for me...)
    - proportion of variation in the outcome explained by the model
- with only 1 predictor (age), neither concordance nor R^2 are very high
- Likelihood ratio test/ Wald test/ Score test
    - null hypothesis testing
    - at small p-value, null is rejected
    - i.e. **model with age in it, is significantly better than null model**

### practice: ethnic groups
```{r}
library(survminer)

g$ethnicgroup <- factor(g$ethnicgroup)

cox_ethnicity <- coxph(Surv(fu_time, death) ~ ethnicgroup, data =g)
summary(cox_ethnicity)

```

- large standard errors, large CIs...
- hazard ratio, need to know reference ethnicity, which is ethnicgroup1 (white)
    - group2 (black) not significantly different hazard
    - group3 (indian) significantly different hazard
    - group9 not significant

#### 43 had missing ethnicities
- make an unknown ethnicity group and rerun
```{r}
g$ethnicgroup %>% levels()
levels(g$ethnicgroup) <- c(levels(g$ethnicgroup), "8")
g$ethnicgroup[is.na(g$ethnicgroup)] <- "8"

cox_ethnicity_noNAs <- coxph(Surv(fu_time, death) ~ ethnicgroup, data = g)
summary(cox_ethnicity_noNAs)
```

- found that including NA ethnicities as a group did not change results by much
- but if including multiple predictors, NAs will add up, will affect results,
- better to try recoding NAs or impute, and compare results before and after


### EOF