---
title: "week4_6final_analysis"
author: "Hippo"
date: "`r format(Sys.time(), '%Y_%m_%d_%H')`"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    fig_width: 4
    fig_height: 3
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
library(tidyverse)
library(ggplot2)
```

### get data
```{r, results="hide"}
library(survival)
library(survminer)
library(car)
g <- read.csv("GMPH_1K_final_2.csv", header = TRUE, sep = ",")
g %>% dim()
```

### predictors from paper
- had: age, gender, ethnicgroup, ihd, stroke, valvular_disease, pvd, copd, pneumonia, hypertension, cancer, metastatic_cancer, senile, dementia, mental_health, prior_dnas, los, cabg
- should have also had: renal_disease, cancer, combined senility and dementia
- shouldn't have had: cabg
<br><br>
- cox model output, should check SE, should be < 1, even better if < 0.1

#### prep predictors
```{r, results="hide"}
# combine senile and dementia into cog_imp, if either senile or dementia = 1, cog_imp = 1
g$cog_imp <- ifelse(g$senile == 1 | g$dementia == 1, 1, 0)
# g$cog_imp %>% table()

# get new df for paper predictors only
g_paper <- g %>% select(c(age, gender, ethnicgroup, ihd, valvular_disease, pvd, stroke, copd, pneumonia, hypertension, renal_disease, cancer, metastatic_cancer, mental_health, cog_imp, los, prior_dnas))

# convert binaries variables to factors
# g_paper %>% str()
g_paper <- g_paper %>% mutate_if(~ n_distinct(.) == 2, as.factor)

# deal with ethnicity NAs
g_paper$ethnicgroup <- as.factor(g_paper$ethnicgroup)
levels(g_paper$ethnicgroup) <- c(levels(g_paper$ethnicgroup), "8")
g_paper$ethnicgroup[is.na(g_paper$ethnicgroup)] <- "8"
# table(g_paper$ethnicgroup, exclude = NULL)

# attach the outcomes
g_paper$fu_time <- g$fu_time
g_paper$death <- g$death

g_paper %>% str()
```

#### continous variables
- check if relationship with outcome is linear
- plots variable (x) vs **Martingale residuals** (y) { why not plot variable vs fu_time ??? }
- do this before cox modeling
```{r}
ggcoxfunctional(Surv(fu_time, death) ~ age + log(age) + sqrt(age), data = g_paper)

ggcoxfunctional(Surv(fu_time, death) ~ los + sqrt(los), data = g_paper) 
# can't log(los), it has some zeros, log(0) = -Inf

ggcoxfunctional(Surv(fu_time, death) ~ prior_dnas + sqrt(prior_dnas), data = g_paper)
# can't log(prior_dnas) neither, same problem
# prior_dnas has a lot of zeros, values are not normally distributed { does it matter??? }
```

#### cox model
```{r}
cox_paper <- coxph(Surv(fu_time, death) ~ ., data = g_paper)
# summary(cox_paper)

HR_allVars <- summary(cox_paper)$coefficients %>%
  as.data.frame() %>% 
  rownames_to_column("vars") %>%
  filter(`Pr(>|z|)` < 0.05) %>%
  select(vars, "HR_allVars" = `exp(coef)`)
HR_allVars
```

- SE are all low enough, would be better if < 0.1 {with larger sample size}

### backward elimination

#### drop variables with p-value > 0.05
```{r}
# get names of only significant variables
sigVars <- summary(cox_paper)$coefficients %>%
  as.data.frame() %>% 
  rownames_to_column("vars") %>%
  filter(`Pr(>|z|)` < 0.05) %>%
  select(vars) %>%
  unlist() %>% unname()
sigVars
```

#### new cox model with significant variables only
- <span style="color:Tomato">somehow course had age, gender, valvular, pneumonia, mets and cog_imp as significant</span>
    - clearly wrong since output from first cox model is the same...
```{r}
g_paper %>% str()
cox_sigVars <- coxph(Surv(fu_time, death) ~ age + gender + valvular_disease + pneumonia + metastatic_cancer + cog_imp, data = g_paper)
HR_sigVars <- summary(cox_sigVars)$coefficients %>%
  as.data.frame() %>%
  rownames_to_column("vars") %>%
  select(vars, "HR_sigVars" = `exp(coef)`)
HR_sigVars
```

#### compare HR of original cox model and model with only significant vars
```{r}
inner_join(HR_allVars, HR_sigVars, by = "vars")
```

- only metastatic_cancer HR had dramatic change (8.98 to 12.20)
    - should we add back variables to try to get metastatic_cancer HR back to 8.98?
    - depends on how results is going to be used,
    - if you only care about finding significant predictors, then HR doesn't really matter
    - { then course notes goes on to look at cog_imp while talking about metastatic_cancer... }
- { skipped assumptions testing, no new info }

### why final model differ from report
1. data similar, not the same
    - data here is simulated from same distribution as original, not same data
2. some variables omitted
3. logistic regression vs survival analysis (different outcomes)
    - logistic regression used mortality within 1 year of admission as outcome
    - survival analysis follow-up period much longer than 1 year
4. size of dataset of original paper much bigger (80,000 vs 1000)
    - smaller dataset = less power to detect significant relations

### EOF