---
title: "week1_3plotKM_survdiff_logRankTest"
author: "Hippo"
date: "`r format(Sys.time(), '%Y_%m_%d_%H')`"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    fig_width: 4
    fig_height: 3
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
library(tidyverse)
library(ggplot2)
```

### what is heart failure
- causes include: coronary heart disease, high blood pressure, arrythmia, alcoholism
- symptoms: shortness of breath, tiredness, swollen ankles
- high mortality and morbidity
- high rate of hospital admission
- readmission high, within 1-4 days of discharge
- hospital admin database gives us data to work with

```{r}
library(survival)
```

### run survfit() and KMplot
```{r, results="hold"}
# read in data
g <- read.csv("GMPH_1K_final_2.csv", header = TRUE, sep = ",")
g %>% dim()

# set variable types
g[,"fu_time"] %>% typeof()
fu_time <- g[,"fu_time"]; fu_time %>% head(20)
g[,"death"] %>% typeof()
death <- g[,"death"]; death %>% head(20)
```

```{r, out.height="50%", out.width="50%"}
# survfit
km_fit <- survfit(Surv(fu_time, death) ~ 1) 
# ~ 1 plots KM estimated probability of survival over time for ALL patients, no comparisons made
plot(km_fit)
```

#### interpret the function
- survfit()
    - fits simple survival model,
    - no predictor given, model has only the intercept (i.e. ~ 1)
- Surv()
    - 2 arguments: follow-up time of each patient + if they died
    - death = 1, people who died by the end of their follow-up period
    - death = 0, people still alive by end of their follow-up period, censored
- survfit() produces KM estimates of "the probabilty of survival over time",
    - used by plot to produce KM curve

#### table of survival estimates
```{r}
summary(km_fit, times = c(1:7, 30, 60, 90*(1:10)))
```

- t=1, survival probability is high (99%)
- t=900, survival probability is low (38%)
- { longer you stay in hospital the lower your survival prob?  
or is this just natural decrease in surv prob over time??? }
<br><br>
- see week1_4 for calculating survfit() output by hand

### run survfit() with groups
```{r, out.height="50%", out.width="50%"}
g[,"gender"] %>% typeof()
gender <- as.factor(g[,"gender"])
km_gender_fit <- survfit(Surv(fu_time, death) ~ gender)
plot(km_gender_fit)
```

#### compare survival using logRankTest
```{r}
survdiff(Surv(fu_time, death) ~ gender, rho = 0)
```

- rho = 0 is default
- yields log-rank (aka Mantel-Haenszel) test
- **log-rank test does NOT have proportional hazard assumption**
- { is log-rank test non-parametric? }
<br><br>
- Observed = weighted observed number of events in each group
- Expected = weighted expected number of events in each group

```{r}
# explore log-rank test
# ?survdiff
# tests if there's a difference btw 2 or more survival curves,
# using rho family of tests

g %>% filter(gender == 1) %>% summarise(., mean(fu_time)) # 460.8942
g %>% filter(gender == 2) %>% summarise(., mean(fu_time)) # 472.0841
```

### log-rank test details
- [https://en.wikipedia.org/wiki/Logrank_test](https://en.wikipedia.org/wiki/Logrank_test)
- compares survival distributions of 2 samples
- nonparametric
- good when data is right skewed and censored
- **logrank test statistic**
    - compares estimates of hazard function of 2 groups at EACH time point
    - constructed by computing O and E number of events in one group at each time point,
        - then sum to obtain overall summary across all-time points where there was an event
    - <span style="color:Tomato">more details needed...</span>
- **null hypothesis**
    - 2 groups have identical hazard functions (survival curves)
    - H0: h_1(t) = h_2(t)
- assumptions
    - censoring is unrelated to hazard
    - survival probability same regardless of when subject was recruited (early vs later in study)

### exercise: compare survival of age groups
- age_65+ vs otherwise
```{r, out.height="50%", out.width="50%"}
age_65plus <- ifelse(g[,"age"] >= 65, 1, 0)
table(age_65plus)
# table(g[, "age"], age_65plus)

km_age65plus_fit <- survfit(Surv(fu_time, death) ~ age_65plus)
plot(km_age65plus_fit)
survdiff(Surv(fu_time, death) ~ age_65plus, rho=0)
```

### EOF