---
title: "week3_2coxModel_that_doesnt_converge"
author: "Hippo"
date: "`r format(Sys.time(), '%Y_%m_%d_%H')`"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
    fig_width: 4
    fig_height: 3
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
library(tidyverse)
library(ggplot2)
```

### run with these predictors
- age, gender, copd, quintile, and ethnic group
- quintile is 5 socioeconomic categories (1 = most affluent, 5 = poorest),
    - nationally weighted by population, meaning that 20% of England's population live in areas of each quintile
```{r}
library(survival)
g <- read.csv("GMPH_1K_final_2.csv", header = TRUE, sep = ",")
g %>% dim()
g %>% head()
g$gender <- as.factor(g$gender)
g$ethnicgroup <- as.factor(g$ethnicgroup)
g$copd <- as.factor(g$copd)

cox_convg <- coxph(Surv(fu_time, death) ~ age + gender + copd + quintile + ethnicgroup, data = g)
summary(cox_convg)
```

### running with quintiles as categorical leads to non-convergence
```{r}
g$quintile <- factor(g$quintile)

cox_nonconvg <- coxph(Surv(fu_time, death) ~ age + gender + copd + quintile + ethnicgroup, data = g)
"Loglik converged before variable  4,5,6,7,8 ; coefficient may be infinite."
summary(cox_nonconvg)
```

### quintile CIs infinitely wide

#### table quintile
```{r}
table(g$quintile, exclude = NULL)
```

- quintile 0 is where postal code was missing, couldn't find socioeconomic status from national file
- regardless there's only 4 in that group,
    - but sometimes 4 is even enough for convergence,
    - problem is deeper

#### table quintile with death
```{r}
round(prop.table(table(g$quintile, g$death),1)*100,1)
```

- real problem is that quintile=0 is the reference group
- and none of the 4 patients were dead
- model had no reference group to compare to

### solutions

#### change reference category
- R sets lowest category as reference by default
- in terms of socioeconomic status, usually interest in effect of lower status vs higher status {lower should be reference}
- so let's set quintile = 1 as reference

#### combine categories
- quintile = 0 are not NAs
- they don't have postal code, might be from overseas, homeless, new immigrants
- but quintile = 0 group so small, combining it with NA and creating a new group shouldn't impact results

#### exclude patients
- if only 4/1000 patients, might be better to just exclude these

#### dropping variable
- if a lot of patients had quintile = 0, doesn't make sense to drop all patients,
- better to drop that variable


### EOF